---
title: "ANOVA Test"
output:
  html_document:
    df_print: paged
---

```{r tidyverse library, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


### Summary
The dataset contains the mean number of years in school among men between ages of 25 and 
34 years old by country and year. The countries for United States, Canada, Mexico, and China will be analysed using a one-way ANOVA hypothesis test. This data set is sourced from Gapminder (https://www.gapminder.org/data/).

### Null and Alternative Hypotheses

Let $\mu_1$ = United States mean years in school men 25 to 34

Let $\mu_2$ = Canada mean years in school men 25 to 34

Let $\mu_3$ = Mexico mean years in school men 25 to 34

Let $\mu_4$ = China mean years in school men 25 to 34  
<br>
$H_0$: $\mu_1$ = $\mu_2$ = $\mu_3$ = $\mu_4$

$H_1$: At least one sample mean is significantly different from the others.


### Read CSV data file
```{r read csv file}
df <- read.csv("data/mean_years_in_school_men_25_to_34_years.csv")

head(df)
```

```{r apply filters to dataframe}
df_clean <- df %>%
  filter(geo=='can' | geo=='chn' | geo=='mex' | geo=='usa') %>%
  filter(time >= 2000)

```

### Show aggregate count by country

```{r aggregate count}
df_clean %>%
  group_by(geo) %>%
  count(name = "count")
```

### Show aggregate mean years in school by country

```{r show mean by geo}
df_clean %>%
  group_by(geo) %>%
  summarize(mean=mean(mean_years_in_school_men_25_to_34_years, na.rm = TRUE))
```

### ANOVA Test 

```{r ANOVA test}
anova <- aov(mean_years_in_school_men_25_to_34_years ~ geo, data = df_clean)
summary(anova)
```

### Summary

Since p-value is less than 0.05, we **<span style="color: red;">can</span>** reject the null hypothesis, 
$H_0$: $\mu_1$ = $\mu_2$ = $\mu_3$ = $\mu_4$. There appears to be evidence suggesting that
mean years in school for men 25 to 34 differs among the United States, Canada, Mexico, and China.
<br>

