---
title: "Non-parametric hypothesis testing"
output:
  html_document:
    df_print: paged
---

```{r tidyverse library, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Chi-Squared Test

### Load data file

The dataset contains the House of Representatives voting results (yes or no) among democrats and republicans from 1984 on many issues. This chi-squared test focuses on voting for the handicap infant proposal.

```{r load data file}
df <- read.csv("data/house-votes-84.data")

colnames(df) <- c("Class Name","handicapped-infants", "water-project-cost-sharing",
"adoption-of-the-budget-resolution",
"physician-fee-freeze",
"el-salvador-aid",
"religious-groups-in-schools",
"anti-satellite-test-ban",
"aid-to-nicaraguan-contras",
"mx-missile",
"immigration",
"synfuels-corporation-cutback",
"education-spending",
"superfund-right-to-sue",
"crime",
"duty-free-exports",
"export-administration-act-south-africa")

head(df)
```
### Chi-Squared test for independence of political affiliation and handicapped-infant issue

There are two populations within the dataset which are house of representatives and the infant handicap issue to be voted on.  The house of representatives population has two categories, Republican and Democrat. The infant handicap issue population has two categories, Yes or a No vote.

**Null Hypothesis:**
The number of yes and no votes for the infant handicap issue are independent of political affiliation (democrat or republic). 

**Alternative Hypothesis:**
The number of yes and no votes for the infant handicap issue are <ins>not</ins> independent of political affiliation (democrat or republican).


$$\alpha = 0.05$$

$$H_0: O_{i1} = E_{i1} \ . . . O_{ik} = E_{ik} \  \ i \in \ \{1...r\} $$
$$H_1: Observations\ (yes/no\ votes) \ \ are \ \ not \ \ independent$$



$$k = 2, r = 2$$

```{r function to convert n and y to 0, 1}
map_n_y <- function(x) {
  if (x=="n") 0 else if (x=="y") 1 else x
}

```

### Filter "?" from handicap infants categorical variable

```{r handicap categorical variable}
df_handicap <- df %>% 
  filter(`handicapped-infants`!="?") %>%
  select("Class Name", "handicapped-infants")
```

### Convert "n" and "y" to zeros and ones

```{r convert categories "n" and "y" to 0 and 1}
df_handicap$`handicapped-infants` <- sapply(df_handicap$`handicapped-infants`, map_n_y)

```

### Setup contingency table

```{r convert column to factors}
df_handicap$`handicapped-infants` <- as.factor(df_handicap$`handicapped-infants`)
summary(df_handicap)
```


```{r setup contingency table}
df_table <- table(df_handicap$`Class Name`, df_handicap$`handicapped-infants`)
colnames(df_table) = c("handicap_infants_no", "handicap_infants_yes")

```


```{r display table}
head(df_table)
```

### Chi-squared test for independence

```{r chi-square test}
chisq.test(df_table)
```

### Summary
The null hypothesis states that the number of "yes/no" votes on the handicap infants proposal are independent of political affiliation (republican or democrat).

The calculated p-value is less than the alpha value of 0.05. So, we can reject the null hypothesis. We can conclude that both republican and democratic voting results on handicap infants are not independent of the issue being voted on. We cannot draw any conclusion on the strength of the difference between republican and democratic groups.



