---
title: "One Sample t-test"
author: "Michael Albers"
output: html_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

### Load 2008 Presidential election data for swing states
```{r load dataset}
df = read.csv("data/2008_swing_states.csv", sep=",")
head(df, 10)
```

### Summary of population
```{r summary}
summary(df)
```
### Theoretical mean
```{r theoretical mean}
mu_param = 325000
```


### Null Hypothesis
There is no significant difference between the theoretical mean and the sample mean of the number of 2008 swing votes.

### Alternative Hypothesis
The sample mean of total swing votes is different from the theoretical mean

### Take small random sample of 25 counties (n < 30)
```{r take small sample}
sample_counties = sample(df$total_votes, 25, replace = FALSE)

```

### Summary of Sample
```{r first sample summary}
summary(sample_counties)
```

### Perform t-test for one-population mean of total votes
```{r t-test}
df_total_votes = na.omit(df$total_votes)
t.test(df_total_votes, mu = mu_param)
```
### Results

The p-value is less than 0.05, so we **<span style="color: red;">can</span>** reject the null hypothesis that the sample mean is equal to the theoretical mean of 325,000 swing votes.



