---
title: "Two Sample t-test"
author: "Michael Albers"
output: html_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

### T-Test for Two Independent Populations

$$let\ \mu_0 = number\ of\ survivors\ in\ first\ class$$
$$let\ \mu_1 = number\ of\ survivors\ in\ third\ class$$


$$H_0:\mu_0 - \mu_1 = 0$$
$$H_1:\mu_0 - \mu_1 \neq 0$$
$$\alpha = 0.05$$

### Load Titanic CSV file into dataframe

```{r load dataframe}
df = read.csv("Titanic.csv", sep = ",")
```

### Preview dataframe
```{r head(df)}
head(df, 10)
```

### Group summary statistics
```{r group summary}
df %>% 
  group_by(Class) %>%
  filter(Survived=="Yes") %>%
  summarize(mean_freq = mean(Freq))
```


### First class passengers
```{r first class passengers}
first_class <- na.omit(filter(df, Class=="1st" & Survived=="Yes"))
```

### Third class passengers
```{r third class passengers}
third_class <- na.omit(filter(df, Class=="3rd" & Survived=="Yes"))
```

### T-test Case 1: homoscedasticity 
$$\sigma_1 = \sigma_2$$

```{r t-test case 1}
t.test(first_class$Freq, third_class$Freq, alternative='two.sided', var.equal=TRUE)
```


### T-test Case 2: heteroscedasticity
$$\sigma_1 \neq \sigma_2$$
```{r t-test case 2}
t.test(first_class$Freq, third_class$Freq, alternative='two.sided', var.equal=FALSE)
```


### Paired T-test for Two Dependent Population Mean difference

"The first- and second- class cabins were much closer to the boat deck (where the lifeboats were located) than third class passengers (see Figure 1) (Frey, 2011, 2015).  First- and second-class cabins were primarily separated by purely social barriers whereas physical gates separated the third-class quarters from other areas of the ship (Levinson, 2012, 151). This of course, made accessing lifeboats much easier for first- and second-class passengers, as time is of the essence on a sinking ship." (Titanic: Explore the disasterous sinking of the R.M.S. Titanic. Retrieved from  https://courses.bowdoin.edu/history-2203-fall-2020-kmoyniha/reflection/)

A pair-wise t-test will be performed next since there appears to be a dependency between first and third class passengers aboard the Titanic.

$$let\ \mu_0 = number\ of\ survivors\ in\ first\ class$$
$$let\ \mu_1 = number\ of\ survivors\ in\ third\ class$$

$$H_0:\mu_0 - \mu_1 = 0$$
$$H_1:\mu_0 - \mu_1 \neq 0$$
$$\alpha = 0.05$$

```{r paired t-test}

t.test(first_class$Freq, third_class$Freq, paired = TRUE)

```

### Summary

Since the p-value is greater than 0.05, the null hypothesis is accepted.  There is no significant difference in mean number of survivors in first and third class cabins aboard the Titanic.


**References:**

Titanic: Explore the disasterous sinking of the R.M.S. Titanic. Retrieved from 
Retrieved from https://courses.bowdoin.edu/history-2203-fall-2020-kmoyniha/reflection/


