
---
title: "Data Visualization Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Select a data file (*.csv or other formats) of your own choosing.
2. Import the data file as a dataframe.
3. Create 3 different visualization/plots based on the data. Some visualizations you may want to make include, but are not limited to, bar charts, histograms, scatter plots, box plots. 

## R Markdown

```{r tidyverse library, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


```{r diamonds}
df = read.csv('diamonds.csv')
```

## Information on dataframe

```{r str df 1, echo=FALSE}
summary(df)
```


## Convert cut, color, clarity to factor data type

```{r convert cut, color, clarity to factor datatype}
df$cut = factor(df$cut, levels = c('Fair', 'Good', 'Very Good', 'Premium', 'Ideal'))
df$color = factor(df$color, levels = c('J', 'I', 'H', 'G', 'F', 'E', 'D'))
df$clarity = factor(df$clarity, levels = c('I1', 'SI2', 'SI1', 'VS2', 'VS1', 'VVS2', 'VVS1', 'IF'))
```



```{r str df 2}
str(df)
```

## Show distribution of prices

```{r}
ggplot(df, aes(price)) + 
  geom_histogram(bins = 30) +
  scale_x_continuous(breaks = c(0, 2500, 5000, 7500, 10000, 12500, 15000, 17500, 20000))
```

## Show distribution carat sizes

```{r}
ggplot(df, aes(carat)) + 
  geom_histogram(bins=30) +
  scale_x_continuous(breaks = c(0.05,  0.5,  1.0,  1.5,  2.0,  2.5, 3.0))
```

## Bar plot on cut categorical variable

```{r}
ggplot(df, aes(cut)) +
  geom_bar()
```

## Bar plot on clarity categorical variable

```{r}
ggplot(df, aes(clarity)) +
  geom_bar()
```

## Bar plot for color categorical variable

```{r}
ggplot(df, aes(color)) +
  geom_bar()
```

## Plot cut and price

```{r}
ggplot(df, aes(x=cut, y=price)) + 
  geom_boxplot()
```

## Plot clarity and price

```{r}
ggplot(df, aes(x=clarity, y=price)) +
  geom_boxplot()
```

## Plot color and price

```{r}
ggplot(df, aes(x=color, y=price)) +
  geom_boxplot()
```

## Plot carat (size of diamond) and its effect on price

```{r}
ggplot(df, aes(x=carat, y=price)) +
  geom_point()
```

---
References

Hadley Wickham & Garrett Grolemund. R for Data Science. Retrieved from  https://r4ds.had.co.nz/exploratory-data-analysis.html

---

